---
title: 'Coursera Reproducible Research: Course Project 2'
author: "Chamika Senanayake"
date: "9/14/2020"
output:
  html_document: default
  pdf_document: default
---

# Exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database

## Synopsis 
This project is attempted as a requirement of the Reproducible Research course which is a part Data Science Specialization by John Hopkings University MOOC via Coursera
Storm Data is an official publication of the National Oceanic and Atmospheric Administration (NOAA) which documents:

1. The occurrence of storms and other significant weather phenomena having sufficient intensity to cause loss of life, injuries, significant property damage, and/or disruption to commerce;

2. Rare, unusual, weather phenomena that generate media attention, such as snow flurries in South Florida or the San Diego coastal area; and

3. Other significant meteorological events, such as record maximum or minimum temperatures or precipitation that occur in connection with another event. 

NCDC receives Storm Data from the National Weather Service. The National Weather service receives their information from a variety of sources, which include but are not limited to: county, state and federal emergency management officials, local law enforcement officials, skywarn spotters, NWS damage surveys, newspaper clipping services, the insurance industry and the general public.

## Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

### Questions
Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report


# Preprocessing of Data

## Loading libraries
Setting up preprocessing environment by loading relevant R libraries are crucial. make sure the relevant R environments has installed following packaged before hand. dplyr,tidyr, ggplot & plyr.
````{r, echo=TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(plyr)
````

## Loading data

the relevant dataset is located at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

following script line would check wether storm dataset is already exists, if not it'll downloaded from the above mentioned link. which will prevent unnecessary data usage.

````{r, echo=TRUE, cache = TRUE}
#Database Loading operation
if(!exists("storm.data")) {
  #downloading operation
  if(!file.exists("StormData.csv.bz2")){
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","StormData.csv.bz2")
    print("Storm database file is downloaded!")
  } else {print("Storm database file is already downloaded!")}
  storm.data <- read.csv("StormData.csv.bz2", header = TRUE) 
  print("storm database loaded!")
} else {print("database already exists!")} 
````
## Examination of the Data set

in the storm.data database there are **`r ncol(storm.data)`** Columns and **`r nrow(storm.data)`** Rows, which can be identified using checking data dimentions.
````{r, echo=TRUE}
dim(storm.data)  #Check Dimentions
names(storm.data) #Check headers
str(storm.data) #Check the structure
head(storm.data) #Check first few lines
tail(storm.data) #Check last few lines
````


## Subsetting the dataset accordance with the Questions [scaling down]

the key variable used for this Assignment is
- EVTYPE : e.g. Toranados, flood..

For Question 1, it refers to variable such as event type & variables related to population health. specifically

1. FATALITIES : Number of fatalities
2. INJURIES : Number of Injuries

For Question 2, variable related to types of events have the greatest economic consequences includes

1.PROPDMG : property damages

2.PROPDMGEXP: Units for Property Damage (magnitudes - K,B,M)

3.CROPDMG: Crop Damage

4.CROPDMGEXP: Units for Crop Damage (magnitudes - K,BM,B)


in order to reduce processing resources the data set will be cropped down to what is needed for analysis to solve the questions of this assignment. and the cropped dataset will be checked as below

````{r, echo=TRUE}
storm <- select(storm.data,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))
dim(storm) #Check Dimentions
names(storm) #Check headers
str(storm) #Check the structure
head(storm) #Check first few lines
tail(storm) #Cheack last few line
````

## Missing Values/NA's resolution

There are `r sum(is.na(storm))` Missing Values /NA's in this dataset


# Data Analysis

## Population health dynamics related to event.

### 1. Fatalities
the event type has to be converted to factor variable. then all variables of EVTYPE & FATALITIES were aggregated. out of that portion top 10 were selected and arranged in decending order as following code
````{r, echo=TRUE}
#Analysis of Fatalities with Event type
storm$EVTYPE<-as.factor(storm$EVTYPE)
aggr.fatalites<-aggregate(FATALITIES ~ EVTYPE, data = storm,  FUN="sum") #aggregates fatalitities
top10.fatalities<-aggr.fatalites[order(-aggr.fatalites$FATALITIES), ][1:10, ] #order top 10
````

the top 10 fatality events are as following table

````{r, echo=FALSE}
top10.fatalities
````


### 2. Injuries

variables of EVTYPE & INJURIES were aggregated. out of that portion top 10 were selected and arranged in decending order as following code
````{r, echo=TRUE}
aggr.injuries<-aggregate(INJURIES ~ EVTYPE, data = storm,  FUN="sum") #aggregates injuries
top10.injuries<-aggr.injuries[order(-aggr.injuries$INJURIES), ][1:10, ] #order top 10
````

the top 10 Injury events are as following table

````{r, echo=FALSE}
top10.injuries
````

Above mentioned both Number of Injuries and Number of fatalities against event type can be plotted in one figure as below.

````{r, echo=TRUE}
#plotting of fatalaty vs Evtype chart
par(mfrow = c(1,2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(top10.injuries$INJURIES, names.arg = top10.injuries$EVTYPE, las = 3, main = "10 Injuries Highest Events", ylab = "Number of Injuries")
barplot(top10.fatalities$FATALITIES, names.arg = top10.fatalities$EVTYPE, las = 3, main = "10 Fatalities Highest Events", ylab = "Number of Fatalities", ylim=c(0,max(top10.injuries$INJURIES)))
dev.copy(png, "event-healthplot.png", width = 480, height = 480)
dev.off()
````

## Economical Consequences related to event.

on observing the data related to economical consequences which are crop damages, and property damages the values are labeled as k, m as in a seperated column of exponentional. in order to manage that, we have to turn the exponential columns of crop/property damage from character to numeric giving a measurable variable. we do mapping of values from plyr package as follows

````{r, echo=TRUE}
#dsiplay unique values & assign its values to the same exponential
#Property Damage
unique(storm$PROPDMGEXP)
storm$PROPDMGEXP <- mapvalues(storm$PROPDMGEXP, from=unique(storm$PROPDMGEXP), to =c(
  10^3,# "K" 
  10^6,# "M" 
  1,# ""  
  10^9,# "B" 
  10^6,# "m" 
  10^0,# "+" 
  10^0,# "0" 
  10^5,# "5" 
  10^6,# "6" 
  10^0,# "?" 
  10^4,# "4" 
  10^2,# "2" 
  10^3,# "3" 
  10^2,# "h" 
  10^7,# "7" 
  10^2,# "H" 
  10^1,# "-" 
  10^1,# "1" 
  10^8# "8"
))
storm$PROPDMGEXP <- as.numeric(as.character(storm$PROPDMGEXP))
storm$PROPDMGTOTAL <- (storm$PROPDMG * storm$PROPDMGEXP)/1000000000
#Crop Damage
storm$CROPDMGEXP<-mapvalues(storm$CROPDMGEXP, from = unique(storm$CROPDMGEXP), to = c(
10^0,  # ""  
10^6,  # "M" 
10^3,  # "K" 
10^6,  # "m" 
10^9,  # "B" 
10^0,  # "?" 
10^0,  # "0" 
10^3,  # "k" 
10^2  # "2"
)
                            )
storm$CROPDMGEXP <- as.numeric(as.character(storm$CROPDMGEXP))
storm$CROPDMGTOTAL <- (storm$CROPDMG * storm$CROPDMGEXP)/1000000000
````


after clarifying the exponential its time to calculate the damages in relation to weather event type. for this we have to calculate the total number of damages per each event type after multiplying each and every damage column with exponential column. 

for Property Damage, we could run the following R code which will generate the top 10 weather events causing highest property damage

````{r,echo=TRUE}
#Calculate property damage & Display top 10 events causing highest property damage
sumPropertyDamage <- aggregate(PROPDMGTOTAL ~ EVTYPE, data = storm,  FUN="sum")
propdmg10Total <- sumPropertyDamage[order(-sumPropertyDamage$PROPDMGTOTAL), ][1:10, ]
propdmg10Total
````

Similarly we could run the following R code which will generate the top 10 weather events causing highest crop damage

````{r,echo=TRUE}
#Calculate crop damage & Display top 10 events causing highest crop damage
sumCropDamage <- aggregate(CROPDMGTOTAL ~ EVTYPE, data = storm,  FUN="sum")
cropdmg10Total <- sumCropDamage[order(-sumCropDamage$CROPDMGTOTAL), ][1:10, ]
cropdmg10Total
````

now Finally we could create a barplot figure comparing both property and crop damage in relation to type of weather event.
````{r,echo=TRUE}
par(mfrow = c(1,2), mar = c(12, 4, 3, 2), mgp = c(3, 1, 0), cex = 0.8)
barplot(cropdmg10Total$CROPDMGTOTAL, names.arg = cropdmg10Total$EVTYPE, las = 2, main = "10 Crop Damages Highest Events", ylab = "Damage Crop Values (in Billions) ", ylim=c(0,max(propdmg10Total$PROPDMGTOTAL)))
barplot(propdmg10Total$PROPDMGTOTAL, names.arg = propdmg10Total$EVTYPE, las = 3, main = "10 Property Damages Highest Events", ylab = "Damage Property Values (in Billions)")
dev.copy(png, "event-economydamage.png", width = 480, height = 480)
dev.off()
````
# Conclusion

based on this large data set analysis. following can be observed

in terms of population health, Injuries are far greater than fatalities in weather event. out of fatalities toranado is the leading cauase of fatality as well as injuries. 

Property Damage is higher than Crop damage as economical consequences out. Flood is the leading cause of property damage while Drought is the leading cause of crop damage.
